\textbf{\sc Towards Deliberative Control of Autonomous Robots}

### 2012-02-06 trunk/europa-basics.tex
updates

% \eu is a general purpose AI planning toolkit developed at NASA Ames.  
% TODO: Talk about NASA missions where \eu has been used.

@@

% \eu is a versatile planning framework.  let's start by looking at some examples of the main problem types that it can tackle.

@@

Arc Consistency and other, more stringent consistency definitions can be
used to efficiently prove that a CP problem is unsatisfiable. However,
they generally cannot prove that a CP problem is satisfiable and
provide specific variable assignments that constitute solutions. For
that, search algorithms like global backtracking \comment{citation
  needed} or local search are used; these use consistency as an
efficient way to prune the search space \comment {citation CP
  Handbook}. In theory, solving a CP problem is NP-Hard
\cite{ghallab04}, but often very efficient in practice using a number
of consistency and search algorithms that are well understood.

@@

Constraint-based Planning and specifically Constraint-Based Attribute
and Interval planning (CAIP) \cite{mus94,frank2003} is intended to
close that gap. On one hand, it takes advantage of a CSP
representation; on the other it uses attributes and intervals to
maintain an explicit representation of the elements relevant to
planning. This makes it easier to write algorithms that search for an
reason abut plans. The main elements of CAIP are:

@@

\subsection{\eu Plan Representation}
\label{sec:europa:pr}

\eu is an implementation of the CAIP framework. 
\comment{once the CAIP section is finished, this section will need an iteration to provide a few pointers back to it}

\textit{Variables} Values that need to be represented to describe the
problem domain and over which we may want to specify constraints. In the Shopping Agent example, the times at which the agent needs to leave or be back, or executes a purchase, are all instances where variables representing time would be used.

\textit{Objects} The things we wish to describe and refer to in a
domain are considered Objects. As in the case with object-oriented
analysis and design, one can seek out the nouns in any domain
description to find likely objects. In the Shopping Agent example, we night consider the Agent itself, Products and Locations all to be objects. Objects have state and behavior. For example, a Shopping Agent can have:
\begin{enumerate}
	\item State: it's location, a bag that contains the products it has already purchased (the bag can in turn be another object), etc.
	\item Behavior: go to a location to look for a product, perform a purchase, return home, etc. 
\end{enumerate}

As in object-oriented design, objects that have similar state and behavior can be described generically in terms of object types or classes. \eu allows the definition of object classes in the same way that it is done in popular object-oriented programming languages. However, to describe state and behavior for the purposes of planning we need a different mechanism, for that we build on the formalism of first order logic as explained below.

\textit{Tokens}

In first order logic, a predicate defines a relation between objects and properties \comment{citation?}. 
In \eu, we define such relations between variables whose domains are sets of
objects and sets of properties to describe state and behavior. For example, we might use a predicate
\texttt{At($a$,$l$)} to indicate that agent
$a$ is at location $l$, or a predicate \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking acting to buy product $p$. Note that $a$ is a variable which may
have a number of possible values in a problem with multiple
agents. Similarly, $l$ and $p$ are variables whose values are the set of
possible positions and products respectively. \eu can be used to create partial plans, where the domains of variables $a$ and $l$ can have more than one possible value in them, or grounded plans, where single values will be
specified for each variable as we saw in the case of the N-queen's problem.

In general, to come up with an executable plan it is not sufficient to state predicates that describe required state or behavior
without also specifying some temporal extent over which each of those predicates hold. Predicates
that are always true can be thought to hold from the beginning to the
end of time. However, in practice, the temporal extent of interest
must be defined with timepoints to represent its start and end. So we
might want to write \texttt{At($a$,$s$,$e$,$l$)} to indicate that the
agent $a$ is at location $p$ from time $s$ to
time $e$. In fact, this pattern of using such predicates to describe
both state and behavior of objects is so prevalent in \eu that we have
introduced a special construct called a Token which has the built in
variables to indicate the object to which the statement principally
applies and the timepoints over which it holds. In \eu, all predicate
instances are Tokens. Also, in the same way that objects are described generically by classes, in \eu Tokens can be described generically by Token Types.

A Token is an instance of a predicate that represents and object's state or behavior and is defined over
a temporal extent. Every token has five built-in variables:

\begin{enumerate}
    \item \textit {start}: The beginning of the temporal extent over which the predicate
is defined.  
    \item \textit {end}: The end of the temporal extent over which the
predicate is defined.  
    \item \textit {duration}: The constraint \textit{start} + \textit{duration} =
\textit{end} is enforced automatically.  
    \item \textit{object}: The set of objects to
which a token might apply. In a grounded plan each Token applies to a
specific Object, reflecting the intuition that we are using Tokens to
describe some aspect of an Object (i.e. its state or behavior) in
time. However, in a partial plan, the commitment to a specific object
may not yet have been made.
    \item \textit{state}: Tokens can be ACTIVE, INACTIVE, MERGED, or REJECTED. The state
variable captures the token's current state and its reachable states
through further restriction.  This is \eu's mechanism to support CAIP's planning approach, see the Token State Model discussion
below for details.
\end{enumerate}

\comment {resume cleanup here}

\textit{Objects (continued)}

As we have noted, Tokens describe some aspect of an Object in
time. Objects thus may have many Tokens in a plan in order to describe
their state and behavior throughout all points in time of the
plan. Within this general framework, we note a few particulars:

Static Facts: Classes without predicates lead to Objects without
Tokens. This arises where an objects state or behavior is independent
of time. For example, a domain may have a set of locations and/or
paths for which there is nothing more to say than that they exist.

Timelines: Often objects in a domain must be described by exactly one
token for every given timepoint in the plan. Such objects are so
common that we provide a special construct in \eu to extend these
semantics to derived classes. Any instances of a class derived from a
Timeline will induce ordering requirements among its tokens in order
to ensure no temporal overlap may occur among them. See below for
further discussion.

Resources: Metric resources, e.g. the energy of a battery or the
capacity of a cargo hold, are objects with an explicit quantitative
state in time and with a circumscribed range of changes that can occur
to impact that state i.e. produce, consume, use, change. These changes
are captured as tokens. Resources are such a common requirement for
\eu users that special constructs are also provided for
them. Instances of classes derived from a Resource will induce
ordering requirements on their Tokens in order to ensure that the
level of the resource remains within specified limits.

Timelines

It may be sufficient to maintain only a partial-order among
tokens. However, it is often the case that tokens represent states and
actions of a single object in the system. Such tokens are typically
mutually-exclusive. \eu uses a Timeline structure developed in (??)
to concisely capture system components whose behavior is described
over time in this manner.

\comment{This is the first instance that this domain is being
  introduced. Either we need more explanation or it ought to be
  removed.} For example, consider the tire-world domain. The tire was
located in the trunk with the predicate tireLocated(Trunk) and located
on the ground with the predicate tireLocated(Ground). Clearly, the
same tire cannot be in both places at once. A Timeline provides a
simple method for aggregating the statements about a tire such that
they are mutually exclusive.

Figure 1: A Timeline for a Tire

Figure 1 illustrates how a Timeline can be used to describe the
whereabouts of a tire. The tire is an object in the tire-world
represented as a timeline. It has predicates associated with it which
can describe its states and actions over time. The predicate names
need not contain the tire prefix; this is now implicit since the
tokens are assigned to a specific tire instance (i.e. an instance of a
Timeline). While Timelines are a useful element of the \eu planning
paradigm, they are not essential. A more general notion of a system
Object can be used which does not impose restrictions of
mutual-exclusion or non-zero duration. This can be important in
supporting partial-order planning.

For this example, the state of the tire is specified using 3
contiguous tokens. The end and start time-points are related by an
equality constraint. A Moving predicate has been introduced to cover
the transition from one location to another. It takes 2 location
arguments. Notice that precedence relationships exist between tokens
such that they cannot overlap but the start and end times may remain
flexible. To illustrate this, sample times are included. Assume a
total time range of interest between 0 and 1000. In addition, tokens
on Timelines have a minimum duration of 1. As a result the values
shown are the most flexible possible values for the start and end of
each token. There are a number of advantages of allowing this
flexibility. First, the basic structure of the plan can be developed
without over-committing to specific times. If it is not necessary for
the tire to be on the ground at time-step 4, then a planner should not
be forced to specify it. Such an approach permits a least-commitment
approach to planning. Second, in many domains it is simply impossible
to know in planning exactly how long an activity might take. For
example, a common activity of driving a car from one point to another
on a road can take varying amounts of time depending on traffic and
traffic lights. In such circumstances it is more practical to express
upper and lower bounds on durations which naturally lead to intervals
for start and end times. In such domains, flexibility in planning aids
robustness in execution.  Rules

In order for a plan to be valid, it must comply with all rules and
regulations pertinent to the application domain in question. Rules
govern the internal and external relationships of a token. For
example, consider a parameterized predicate describing a transition
from one location to another. Let the parameters be from and to. The
parameters are instantiated on a token as variables whose domain of
values is the set of all locations in a given problem. A rule
governing an internal relationship among token variables might
stipulate that a transition must involve a change in location. This
can be easily expressed as a constraint on the definition of a
predicate of the form: from != to. It is reasonable to further
stipulate that one must be Located somewhere before a transition can
occur, and one must end up Located somewhere when completed. This is
an example of a rule governing an external relationship among
tokens. It specifies a requirement that tokens of the predicate
Located precede and succeed tokens of the predicate Moving.

Figure 2: Internal and external relationships from rules on Moving
Figure 2 illustrates the entities and relations involved in specifying
such a rule on a Moving predicate. The token on which the rule applies
is referred to as the master. Each Located token required by the
master is referred to as a slave. All variables are indicated by name
and their domains are expressed as intervals in the case of temporal
variables and as enumerations for the remainder. Application of a rule
on a token can thus cause slave tokens, variables, and constraints to
be introduced.  Token State Model

The capability of a domain rule to cause a slave token to be created
is a key vehicle through which planning occurs. Semantically, this
rule imposes a requirement for supporting tokens to be in the plan in
order for the master to be valid. There are 2 possibilities to
consider:

The slave is inserted as an active token in the plan. As such, rules
may be activated on the slave, and it may consume resources.

The token is merged with a matching token already in the plan. Once a
slave is merged, the requirement it represents are considered
satisfied. The process of merging passes on all restrictions imposed
on the slave to the active token upon which it is merged. Merging a
token requires finding a target active token that is compatible with
the inactive token. For an active token and an inactive token to be
compatible requires that they are instances of the same predicate and
that no intersections between corresponding variables are empty. The
effect of merging is illustrated in Figure 3.

Figure 3: Merging an inactive token on an active token. Domain
restrictions occur in highlighted variables of active token.

On creation of a token, where the commitment has not been made yet to
activate or merge the token, the token is said to be inactive. There
is a nuance to the state model for tokens which relate to the mode of
its creation. If the token is allocated explicitly by an external
actor, rather than internally through rule firing, it may include the
state rejected indicating the planner is permitted to reject the
token. A valid plan can include rejected tokens. This typically arises
where the token represents a goal that is preferable to achieve but
not mandatory. This state is not reachable for slaves since that would
imply selective adherence to the domain model. Figure 4 presents the
state transition diagram for token states relevant for planning. The
transitions are operations on a partial plan which can decide an
outcome for an inactive token. As operations which may arise in
search, they must be reversible during backtracking. The cancellation
operations in each case are also shown.

Figure 4: Token States and Transitions for Planning

The state of a token is embodied with a 5th and final built-in
variable referred to as the state variable. The planner states are
values in the domain of this variable {MERGED, ACTIVE, REJECTED}. The
INACTIVE state is captured by the variable being unbound. If a value
is removed from the domain, that state will not be reachable. For
example, when a slave is created via a rule, the REJECTED value is
removed from its state variable.

\textit{Summary}

This has presented the main elements of the EUROPA planning approach. They are:

\item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
states and actions in time. Note that tokens do not discriminate
between state and action.

\item Token States which are the basis of planning operations on a partial
plan.

\item Constraints to describe relationships among tokens. This provides an
expressive method of describing interactions among tokens in the
context of Temporal Planning.

\item Timelines as a concise abstraction to express the evolution of state
and behavior for a system component. It provides semantics of mutual
exclusion sequencing in time. Other core abstractions are available in
\eu for handling metric resources.

\item Domain rules to describe internal and external relationships on and
between tokens respectively. Rules are applied to active tokens
(master tokens) referred to as masters and typically produce inactive
tokens (slaves) which are a key vehicle through which the planning
process evolves. The rule structure in \eu differs from the more
restrictive commitment to preconditions and effects in classical
planning which prohibits durative actions and disembodied effects
(i.e. effects which may not occur until some temporal distance after
the end of the action).

>>>>

@@

@@

Arc Consistency and other, more stringent consistency definitions can
be used to efficiently prove that a CP problem is
unsatisfiable. However, they generally cannot prove that a CP problem
is satisfiable and provide specific variable assignments that
constitute solutions. For that, search algorithms like global
backtracking \comment{citation needed} or local search are used; these
use consistency as an efficient way to prune the search space
\cite{cp06}. In theory, solving a CP problem is NP-Hard
\cite{ghallab04}, but often very efficient in practice using a number
of consistency and search algorithms that are well understood.

@@

Constraint-based Planning and specifically Constraint-Based Attribute
and Interval planning (CAIP) \cite{mus94,frank2003} is intended to
close that gap. On one hand, it takes advantage of a CSP
representation; on the other it uses attributes and intervals to
maintain an explicit representation of the elements relevant to
planning. \eu is an implementation of the CAIP framework.  This makes
it easier to write algorithms that search for an reason abut
plans. The main elements of CAIP are:

@@

### 2012-02-06 trunk/europa.tex
updates

The following figure displays the main architectural components in EUROPA and their relationships:

\begin{figure}
\centering
\includegraphics[scale=0.5]{figs/EUROPA-Architecture.jpg}
\caption{\small EUROPA's Architecture}
\label{fig:europa-architecture}
\vskip+0.1cm
\end{figure}

@@

	\item The Constraint Reasoning Engine (CRE): manages variables, the domains from which they can take values, and constraints that define relationships among variables. It also provides an efficient arc consistency mechanism (TODO: ref to AC-3). The CRE is designed so that specialized reasoning algorithms for specific constraints can be easily and efficiently plugged in.

	\item The Plan Database (PDB): manages object types and objects, which are a mechanism to group variables to more naturally model the real world in much the same way that object oriented programming does. Also manages tokens, which are a mechanism to group variables to represent temporally scoped state. Objects and tokens can be used to model planning domains in a much more natural and extensible way that a pure CSP approach can.

	\item The Rules Engine (RE): manages token relationships in a planning model

	\item The language modules: NDDL, ANML, and other higher level modeling languages can me used to create domain models and problem instances (TODO: point advantages over pure programming API approaches, provide snippets)

	\item API Layer: makes all the services available so that EUROPA can be used to build applications.

	\item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
		\begin{enumerate}
			\item Temporal Network
			\item Resources
			\item NDDL
			\item ANML (TODO:maybe it's better to leave out?)
		\end{enumerate}

@@

EUROPA can ingest descriptions of models, plans and goals. This means that a single planner can be applied to different systems and problems, simply by providing different models and goals. For such planners the expressiveness of the languages for models, goals and plans is of great importance. EUROPA provides a high level language called NDDL (New Domain Description Language) that allows the user to specify the different components of a problem domain in a concise way. The main characteristics that make NDDL a powerful tool for describing problem domains are:

\begin{enumerate}
	\item It's Object Oriented: NDDL supports classes and inheritance in similar fashion to popular Object-Oriented languages like Java and C++. It also supports polymorphism for some of the planning components. Using object classes and instances is a time-tested approach to naturally describe problem domains (TODO: OO ref?).
	\item It offers declarative constructs to define constraints and causality dependencies between actions and effects.
	\item It offers procedural constructs to populate EUROPA's plan database with specific problem instances expressed in terms of objects, variables, constraints, facts and goals.
\end{enumerate}

@@

  starts(effect Own purchase);
  eq(purchase.product,product);
}

@@

TODO: explain variable, object, predicate and action types, including constraints that can be part of them
\begin{enumerate}
	\item Variable Types: This model uses integer (activity duration, start and end times, etc) and string (Location and Product names, etc)
	\item Object Types:
	\item Predicate Types:
	\item Action Types:
\end{enumerate}

Now let's look at how a specific problem instance would be specified in NDDL:

@@

TODO:  explain variable, object, predicate and constraint instances.
explain goals and facts.

@@

TODO: explain how inference is used to detect perform bounds propagation, domain reduction and to detect constraint violations.
- talk about AC3 implementation and extension mechanisms
- talk about europa's constraint library

The Temporal Network:
The notion of time is central to temporal planning. EUROPA uses variables to explicitly represent timepoints for plan activities and states. Constraints among timepoints provide a natural way to express domain axioms. For example, in order to state that activity A must occur before activity B we can say that the end timepoint of A is <= the start timepoint of B.

@@

Dechter et al. (Dechter 1991) proposed that constraints among timepoints can be grouped together to form a Simple Temporal Network (STN). Such a network can be transformed into a Distance Graph (DG) where the outward arc from a node represents the maximum distance from the source node to the target node. The diagram below illustrates a simple STN with just 2 variables and a single constraint. It also shows the resulting DG.

TODO: add figure

Dechter et al. (Dechter 1991) also showed that shortest path algorithms could be used to propagate values in the network and discover a negative cycle. A negative cycle is a path from a node to itself that has a path length less than 0. If such a cycle exists, the network is inconsistent. It was further shown that a single-source shortest path algorithm was sufficient to detect a negative cycle and provide sufficient propagation to yield a backtrack-free search. Thus we have an efficient and complete algorithm for propagating an STN. These results build on the already established notion of a CSP and are naturally incorporated into the general representation and propagation scheme used in EUROPA.

@@

    Reservoir resources can have both 'consume' and 'produce' tokens. Note that these tokens do NOT have the usual start/end/duration token variables, since occur at a single instant in time. Instead, they have a single variable, 'time' used to represent that instant.
    Reusable resources have 'uses' tokens that use a quantity for the duration of the token. These tokens do have the usual start/end/duration variables.
    Unary resources are reusable resources with unit quantity and have 'use' tokens (again, with the usual start/end/duration variables).

    TODO: talk about resource envelope computation by using MaxFlow formulation.

@@

\item A domain model that describes the variable, object, predicate and action types that are relevant for the problem.
\item A problem instance (also called initial state) that consists of:
	\item Variable and Object instances that exist for the entire planning horizon.
	\item Temporally scoped predicate and action instances.
	\item Temporally scoped goals.

@@

All this information is kept in EUROPA's  Plan Database so that inference and search mechanisms can be used to look for a problem solution.

@@

EUROPA provides a built-in solver that performs Plan Space Planning (TODO: ref), in this approach, the initial state is considered a partial plan that needs to be refined toward a solution plan that achieves the goals. The operations to refine the partial plan PP at any time are:
\begin{enumerate}
	\item Find the flaws of PP, that is the conditions that prevent it from being a solution plan (flaw types are explained below).
	\item Select one such flaw
	\item Select a resolver for the flaw
	\item Refine PP by applying the resolver
	\item If an inconsistency is found, try another resolver.
	\item If all possible resolvers for a particular flaw fail, return failure, otherwise continue until resolving all the flaws.
\end{enumerate}

This Plan Space Planning algorithm is translated into EUROPA's representation as follows:

The initial partial plan is the state of EUROPA's Plan Database after the initial state has been instantiated, this results in a set of variable, object and token instances. Inference then takes place to detect flaws in the partial plan, there are 3 kinds of flaws that can be detected:

\begin{enumerate}
	\item Unbound Variable: a variable in the partial plan whose domain is not a singleton. Unbound Variables are resolved by specification of a value from the domain of the variable.

	\item Open Condition: an open condition is an inactive token. Inactive tokens can be generated by explicitly posted goals, or when a token is activated, rules may fire that create inactive slave tokens (TODO: point to more detailed decription above in Rules Engine section).  Open Conditions can be resolved in three ways: 
\begin{enumerate}
		\item Merging: A free activity is merged with a matching activity already in the plan. An activity a is said to match an activity a? if a and a? unify and the temporal constraints involving a are satisÞed by a?. Thus,  a and a? can be considered the same activity and we do not need to introduce a in the plan. Consequently, the compatibilities associated with a are not Þred, because they have been already triggered when a? was introduced in the plan. 
		\item Activation: We introduce a new activity a in the current plan associating it with the proper timeline, but without choosing a speciÞc time slot for it.The compatibilities associated with a are applied and the subgoal activities resulting from those compatibilities are introduced as free activities. This results in both an ordering ßaw, corresponding to the just activated activity, and a number of open condition ßaws, corresponding to the new subgoal activities. 
		\item Rejection: Some goals may be optional, if that is the case for a particular order condition, its corresponding flaw can be resolved by discarding the inactive token.
\end{enumerate}

@@

The Plan Space Planning algorithm described above can be implemented in many different ways depending on the approach chosen for flaw and resolver selection and for backtracking. EUROPA's built-in solver implements a chronological backtracking algorithm that is summarized in Algorithm \ref{alg:europa:solve}.
 % summarized in the figure below.

@@

TODO: expand a little more on how heuristics can be implemented by applying ordering/filtering for chooseFlaw and makeDecisionPoint
examples:
- deal with earliest flaws first
- deal with planning (open condition flaws) before scheduling (threats) ones

@@

TODO: 
- explain how other solvers can be built, maybe use NQueens or RCPSP example to illustrate. Or maybe build a little specialized solver for shopping or rover examples.

>>>>

\begin{figure}
\centering
\includegraphics[scale=0.5]{figs/EUROPA-Architecture.jpg}
\caption{\small The \eu Architecture}
\label{fig:europa-architecture}
\vskip+0.1cm
\end{figure}

Fig. \ref{fig:europa-architecture} shows the main architectural
components in \eu and their relationships.

The Constraint Reasoning Engine (CRE) manages variables, the domains
from which they can take values, and constraints that define
relationships among variables. It also provides an efficient arc
consistency mechanism \cite{mackworth77}. The CRE is designed so that
specialized reasoning algorithms for specific constraints can be easily
and efficiently plugged in. The Plan Database (PDB) manages object types
and objects, which are a mechanism to group variables to more naturally
model the real world in much the same way that object oriented
programming does. Also manages tokens, which are a mechanism to group
variables to represent temporally scoped state. Objects and tokens can
be used to model planning domains in a much more natural and extensible
way that a pure CSP approach can. The Rules Engine (RE) manages token
relationships in a planning model. The language modules (NDDL, ANML, and
other higher level modeling languages) are used to create domain models
and problem instances \comment{point advantages over pure programming
  API approaches, provide snippets}. Finally the API Layer (Fig.
\ref{fig:europa-api}) makes all the services available so that \eu can
be used to build applications.

\comment{Is Fig. \ref{fig:europa-api} necessary?}

	% \item Extension Modules bundled with EUROPA: Temporal Network, Resources, NDDL, ANML.
	% 	\begin{enumerate}
	% 		\item Temporal Network
	% 		\item Resources
	% 		\item NDDL
	% 		\item ANML (TODO:maybe it's better to leave out?)
	% 	\end{enumerate}

\eu can ingest descriptions of models, plans and goals and thereby can
be applied to different systems and problems, simply by providing
different models and goals. Consequently the expressiveness of the
languages for models, goals and plans is of great importance. For \eu
NDDL (New Domain Description Language [pronounced ``noodle'']
\comment{cite}) allows the user to specify the different components of a
problem domain in a concise way. The main characteristics that make NDDL
a powerful tool for describing problem domains are:

@@

\item Object Oriented: NDDL supports classes and
  inheritance in similar fashion to popular Object-Oriented languages
  like Java and C++. It also supports polymorphism for some of the
  planning components. Using object classes and instances is a
  time-tested approach to naturally describe problem domains \comment{OO
    citation}.

\item It offers declarative constructs to define constraints
  and causality dependencies between actions and effects.

\item It offers procedural constructs to populate \eus plan database
  with specific problem instances expressed in terms of objects,
  variables, constraints, facts and goals.

@@

@@

  starts(effect Own purchase);
  eq(purchase.product,product);
}
\end{verbatim}

@@

\comment{explain variable, object, predicate and action types, including constraints that can be part of them}
\begin{enumerate}
	\item Variable Types: This model uses integer (activity duration, start and end times, etc) and string (Location and Product names, etc)
	\item Object Types:
	\item Predicate Types:
	\item Action Types:
\end{enumerate}

 shows how a specific problem instance would be
specified in NDDL.

@@

\label{eu:model}
\end{verbatim}

\comment{explain variable, object, predicate and constraint instances.
explain goals and facts.}

@@

\comment{explain how inference is used to detect perform bounds
  propagation, domain reduction and to detect constraint violations.} 
- talk about AC3 implementation and extension mechanisms
- talk about europa's constraint library

The Temporal Network: The notion of time is central to temporal
planning. \eu uses variables to explicitly represent timepoints for plan
activities and states. Constraints among timepoints provide a natural
way to express domain axioms. For example, in order to state that
activity A must occur before activity B we can say that the end
timepoint of A is $leq$ the start timepoint of B.

@@

\cite{dechter91} proposed that constraints among timepoints can be
grouped together to form a Simple Temporal Network (STN). Such a network
can be transformed into a Distance Graph (DG) where the outward arc from
a node represents the maximum distance from the source node to the
target node. They also showed that shortest path algorithms could be
used to propagate values in the network and discover a negative cycle. A
negative cycle is a path from a node to itself that has a path length
less than 0. If such a cycle exists, the network is inconsistent. It was
further shown that a single-source shortest path algorithm was
sufficient to detect a negative cycle and provide sufficient propagation
to yield a backtrack-free search. Thus we have an efficient and complete
algorithm for propagating an STN. These results build on the already
established notion of a CSP and are naturally incorporated into the
general representation and propagation scheme used in \eu.
The diagram below illustrates a simple STN with just 2
variables and a single constraint. It also shows the resulting DG.

\comment{ add figure}

@@

Reservoir resources can have both 'consume' and 'produce' tokens. Note
that these tokens do NOT have the usual start/end/duration token
variables, since occur at a single instant in time. Instead, they have a
single variable, 'time' used to represent that instant.

Reusable resources have 'uses' tokens that use a quantity for the
duration of the token. These tokens do have the usual start/end/duration
variables.

Unary resources are reusable resources with unit quantity and have 'use'
tokens (again, with the usual start/end/duration variables).

\comment{talk about resource envelope computation by using MaxFlow formulation.}

@@

\item A domain model that describes the variable, object, predicate and
  action types that are relevant for the problem.

\item A problem instance (also called initial state) that consists of:
	\item Variable and Object instances that exist for the entire planning horizon.
	\item Temporally scoped predicate and action instances.
	\item Temporally scoped goals.

@@

All this information is kept in \eus Plan Database so that inference and
search mechanisms can be used to look for a problem solution.

@@

\eu provides a built-in solver that performs Plan Space Planning
\cite{ghallab04} in this approach, the initial state is considered a
partial plan that needs to be refined toward a solution plan that
achieves the goals. The operations to refine the partial plan PP at any
time are:

\begin{enumerate}
	\item Find the flaws of PP, that is the conditions that prevent it from being a solution plan (flaw types are explained below).
	\item Select one such flaw
	\item Select a resolver for the flaw
	\item Refine PP by applying the resolver
	\item If an inconsistency is found, try another resolver.
	\item If all possible resolvers for a particular flaw fail, return failure, otherwise continue until resolving all the flaws.
\end{enumerate}

This Plan Space Planning algorithm is translated into \eus
representation as follows:

The initial partial plan is the state of \eus Plan Database after the
initial state has been instantiated, this results in a set of variable,
object and token instances. Inference then takes place to detect flaws
in the partial plan, there are 3 kinds of flaws that can be detected:

\item Unbound Variable: a variable in the partial plan whose domain is
  not a singleton. Unbound Variables are resolved by specification of a
  value from the domain of the variable.

\item Open Condition: an open condition is an inactive token. Inactive
  tokens can be generated by explicitly posted goals, or when a token is
  activated, rules may fire that create inactive slave tokens (TODO:
  point to more detailed decription above in Rules Engine section). Open
  Conditions can be resolved in three ways:

  \item Merging: A free activity is merged with a matching activity
    already in the plan. An activity a is said to match an activity a?
    if a and a? unify and the temporal constraints involving a are
    satisÞed by a?. Thus, a and a? can be considered the same activity
    and we do not need to introduce a in the plan. Consequently, the
    compatibilities associated with a are not Þred, because they have
    been already triggered when a? was introduced in the plan.

  \item Activation: We introduce a new activity a in the current plan
    associating it with the proper timeline, but without choosing a
    speciÞc time slot for it.The compatibilities associated with a are
    applied and the subgoal activities resulting from those
    compatibilities are introduced as free activities. This results in
    both an ordering flaw, corresponding to the just activated activity,
    and a number of open condition flaws, corresponding to the new
    subgoal activities.

  \item Rejection: Some goals may be optional, if that is the case for a
    particular order condition, its corresponding flaw can be resolved
    by discarding the inactive token.

@@

The Plan Space Planning algorithm described above can be implemented in
many different ways depending on the approach chosen for flaw and
resolver selection and for backtracking. \eus built-in solver implements
a chronological backtracking algorithm that is summarized in Algorithm
\ref{alg:europa:solve}.

@@

\comment{expand a little more on how heuristics can be implemented by applying ordering/filtering for chooseFlaw and makeDecisionPoint}
examples:
- deal with earliest flaws first
- deal with planning (open condition flaws) before scheduling (threats) ones

@@

\comment{explain how other solvers can be built, maybe use NQueens or
  RCPSP example to illustrate. Or maybe build a little specialized
  solver for shopping or rover examples.}
